Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on

### 2011-04-18 paper.tex
Added more text about CDK taverna. Added a section about databases. Added rows to BO table and added subheadings to table.

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
The Blue Obelisk movement was established in 2005 at the
229\textsuperscript{th} National Meeting of the American Chemistry
Society as a response to the lack of open data, open standards and
open source (ODOSOS) in chemistry. While other scientific disciplines
such as physics, biology and astronomy (to name a few) were embracing
new ways of doing science and reaping the benefits of community
efforts, there was little if any innovation in the field of chemistry
and scientific progress was actively hampered by the lack of access to
data and tools.

Since 2005 it has become evident that a good amount of development in open 
chemical information and information is driven by the demands of neighbouring
scientific fields. In many areas in biology, for example, the importance of 
small molecules and their interactions and reactions in biological systems
has been realised. In fact, one of the first free and open databases and ontologies 
of small molecules was created as a resource about chemical structure and nomenclature
by biologists \cite{DeMatos:2009p3839}.
Open workflow systems such as Taverna (XXX add citation) have opened the door for 
an easy and seamless integration of FOS components into executable workflows 
or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
reported in 2010 \cite{Kuhn:2010p4001}.

@@

@@

Databases?

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Blue Obelisk Open Source software projects}
    (Description if necessary XXXXXXXXXXXXXXX. Add citations to project names.)
    \par \mbox{}
    \par
    \mbox{
      \begin{tabular}{|c|c|c|}
        %% \hline \multicolumn{3}{|c|}{My Table}\\ \hline
        \hline Name & Website & Description? or Lead Developer? \\ \hline
        \multicolumn{3}{|c|}{Cheminformatics toolkits} \\ \hline
        Chemistry Development Kit & http://cdk.sf.net & XXXX  \\ \hline
        Cinfony & http://cinfony.googlecode.com & Noel O'Boyle, Python interface to toolkits \\ \hline
        Indigo & http://ggasoftware.com/opensource/indigo & GGA Software \\ \hline
        Open Babel & http://openbabel.org & Geoffrey Hutchison et al \\ \hline
        RDKit & http://rdkit.org & Greg Landrum \\ \hline
        \multicolumn{3}{|c|}{Interconversion} \\ \hline
        OSRA & http://osra.sf.net & Igor Filippov, Image to structure \\ \hline
        A3 & ..  & .  \\ \hline
      \end{tabular}
      }
  \subsection*{Table 2 - Sample table title}
    Large tables are attached as separate files but should
    still be described here.

>>>>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
The Blue Obelisk movement was established in 2005 at the
229\textsuperscript{th} National Meeting of the American Chemistry
Society as a response to the lack of open data, open standards and
open source (ODOSOS) in chemistry. While other scientific disciplines
such as physics, biology and astronomy (to name a few) were embracing
new ways of doing science and reaping the benefits of community
efforts, there was little if any innovation in the field of chemistry
and scientific progress was actively hampered by the lack of access to
data and tools.
Since 2005 it has become evident that a good amount of development in open 
chemical information and information is driven by the demands of neighbouring
scientific fields. In many areas in biology, for example, the importance of 
small molecules and their interactions and reactions in biological systems
has been realised. In fact, one of the first free and open databases and ontologies 
of small molecules was created as a resource about chemical structure and nomenclature
by biologists \cite{DeMatos:2009p3839}.
Open workflow systems such as Taverna \cite{Hull:2006p60} have opened the door for 
an easy and seamless integration of free and open software (FOS) components into executable workflows 
or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
reported in 2010 \cite{Kuhn:2010p4001}.

@@

CDK-Taverna \cite{Kuhn:2010p4001} as an open-source cheminformatics workflow solution.
Commonly used in this context are workflow engines for cheminformatics, where numerous recurring tasks can be automated, including tasks for chemical data filtering, transformation, curation and migration workflows, chemical documentation and information retrieval related workflows (structures, reactions, pharma- cophores, object relational data etc.) or data analysis workflows (statistics and clustering/ machine learning for QSAR, diversity analysis etc. 
CDK-Taverna has matured to become a freely available and increasingly powerful tool for the biosciences. It was recently ported to 
Taverna version 2.x.
The combination of CDK-Taverna workers with the multitude of workers already published on myexperiment.org by the large and active user community of Taverna enables scientists to quickly build workflows to process a diverse set of data as typically found in today's systems biology scenarios.

@@

\section*{Chemical Structure Registration Systems, Databases}
Databases? (XXX CS: A couple of database-related paragraphs could go here XXX)

XXX the paragraph below on OrChem needs some work
Registration, indexing and searching of chemical structures in relational databases is one of the core areas of cheminformatics. 
A number of structure registration systems have been published in the last five years, exploiting the fact that 
free cheminformatics toolkits such as OpenBabel and the CDK were available. 
OrChem, for example, is an extension for the Oracle 11G database that adds registration and indexing of chemical structures to support fast substructure and similarity searching. The cheminformatics functionality is provided by the Chemistry Development Kit. OrChem provides similarity searching with response times in the order of seconds for databases with millions of compounds, depending on a given similarity cut-off. For substructure searching, it can make use of multiple processor cores on today's powerful database servers to provide fast response times in equally large data sets.
OrChem is free software and can be redistributed and/or modified under the terms of the GNU Lesser General Public License as published by the Free Software Foundation. All software is available via http://orchem.sourceforge.net.

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Blue Obelisk Open Source software projects}
    (Description if necessary XXXXXXXXXXXXXXX. Add citations to project names.)
    \par \mbox{}
    \par
    \mbox{
      \begin{tabular}{|c|c|c|}
        %% \hline \multicolumn{3}{|c|}{My Table}\\ \hline
        \hline Name & Website & Description? or Lead Developer? \\ \hline
        \multicolumn{3}{|c|}{Cheminformatics toolkits} \\ \hline
        Chemistry Development Kit (CDK) \cite{Steinbeck2003, Steinbeck2006} & http://cdk.sf.net & XXXX  \\ \hline
        Cinfony & http://cinfony.googlecode.com & Noel O'Boyle, Python interface to toolkits \\ \hline
        Indigo & http://ggasoftware.com/opensource/indigo & GGA Software \\ \hline
        Open Babel & http://openbabel.org & Geoffrey Hutchison et al \\ \hline
        RDKit & http://rdkit.org & Greg Landrum \\ \hline
        \multicolumn{3}{|c|}{Integration} \\ \hline
        CDK-Taverna \cite{Kuhn:2010p4001} & http://cdk-taverna.blah.XXX & Christoph Steinbeck, Workflow \\ \hline
        A3 & ..  & .  \\ \hline
        \multicolumn{3}{|c|}{Interconversion} \\ \hline
        OSRA & http://osra.sf.net & Igor Filippov, Image to structure \\ \hline
        A3 & ..  & .  \\ \hline
        \multicolumn{3}{|c|}{Structure Databases} \\ \hline
        OrChem \cite{RijnbeekS10} & http://orchem.sourceforge.net & Christoph Steinbeck, Oracle-based Chemical Database Engine \\ \hline
        MyChem  & http://XXX.sourceforge.net & Jerome Pansanel, MySQL-based Chemical Database Engine \\ \hline
        PGChem  & http://XXX.sourceforge.net & Ernst-Georg Schmid, PostgreSQL-based Chemical Database Engine \\ \hline
        Bingo  & http://XXX.XXX.XXX & Dmitry Pavlov, Oracle-based Chemical Database Engine \\ \hline

      \end{tabular}
      }
  \subsection*{Table 2 - Sample table title}
    Large tables are attached as separate files but should
    still be described here.
